---
title: "Predict the Manner of Exercise - Machine Learning Project"
author: "Mani Khanuja"
date: "27 September 2015"
output: html_document
---
#### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. These six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 

+ Class A: exactly according to the specification 
+ Class B: throwing the elbows to the front
+ Class C: lifting the dumbbell only halfway
+ Class D: lowering the dumbbell only halfway 
+ Class E: throwing the hips to the front

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience.
#### Data

The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

#### Objective
The goal of your project is to predict the manner in which subjects did the exercise. This is the “classe” variable in the training set. 
Following needs to performed: 
 + Build a machine learning algorithm to predict activity quality from activity monitors. 
 + Describe the expected out of sample error and estimate the error appropriately with cross-validation
 + Explain the choices made for selecting the machine learning algorithm.

#### Strategy

##### 1. How the model was built
Our outcome variable is classe, a factor variable with 5 levels, as explained above in the background section. 
Three models will be tested using decision tree(rpart), random forest(rf) and tree boosting(gbm) algorithms. The model with the highest accuracy will be chosen as our final model, this might compromise the performance of the code. 

##### 2. Cross Validation
Divided the training data into two data sets. The first one is a training set with 75% of the data which is used to train the model. The second is a validation data set used to assess model accuracy and performance.

##### 3. Expected out-of-sample error
The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Expected accuracy is the accuracy expected in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

#### Code and Conclusion

##### *Load the desired packages*

```{r}
library(caret)
library(randomForest)
library(gbm)
library(plyr)
library(dplyr)
library(AppliedPredictiveModeling)
```

##### *Setting the overall seed for reproduceability*

``` {r}
set.seed(1234)
```

##### *Loading and cleaning data*

+ Load data sets into R
+ Make sure missing values are correctly coded.
+ Removing the columns which are not required.

```{r}
# Read Data from csv downloaded
train = read.csv("/Users/manikhanuja/Documents/github/Machine_Learning/Data/pml-training.csv", na.strings=c("NA","#DIV/0!", "")) 
test = read.csv("/Users/manikhanuja/Documents/github/Machine_Learning/Data/pml-testing.csv", na.strings=c("NA","#DIV/0!", "")) 

# Cleaning data - removing columns which are not required.
training = train[,grepl("belt|arm|dumbbell|classe", colnames(train))]
testing = test[,grepl("belt|arm|dumbbell|classe", colnames(test))]

# Delete columns with all missing values
training<-training[,colSums(is.na(training)) == 0]
testing<-testing[,colSums(is.na(testing)) == 0]
```

##### *Subsetting training data for Cross-Validation*
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r}
# Subsampling data from training set for Cross Validation
inTrain <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subTraining <- training[inTrain, ] 
subTesting <- training[-inTrain, ]
dim(subTraining)
dim(subTesting)
```

##### *Training the Models*
Decision tree, random forest and boosting algorithms are known for their ability of detecting the features that are important for classification. Feature selection is inherent, so it is not so necessary at the data preparation phase.Thus, there won’t be any feature selection section for the models used.

###### **First Alogrithm: Decision Tree**

```{r}
modFitDT <- train(classe ~., method = "rpart", data = subTraining)
print(modFitDT$finalModel)
confusionMatrix(subTesting$classe, predict(modFitDT, subTesting))
```

###### **Second Algorithm: Gradient Boosting Machines**
Gradient Boosted Trees algorithm additionally tries to find optimal linear combination of trees (assume final model is the weighted sum of predictions of individual trees) in relation to given train data. 

```{r}
modFit <- train(classe ~., method = "gbm", data = subTraining, verbose = FALSE)
print(modFit$finalModel)
confusionMatrix(subTesting$classe, predict(modFit, subTesting))
```

##### **Third Algorithm: Random Forests**
Random Forests are trained with random sample of data (even more randomized cases available like feature randomization) and it trusts randomization to have better generalization performance on out of train set.

```{r}
modFitRF <- train(classe ~., method = "rf", data = subTraining)
print(modFitRF$finalModel)
confusionMatrix(subTesting$classe, predict(modFitRF, subTesting))
qplot(predict(modFitRF, subTesting), classe, data = subTesting )
```

##### Conclusion

As expected, Random Forest and boosting algorithm performed better than Decision Trees.

Accuracy for various algorithms used: 
+ Random Forest model: 0.9937 (95% CI: (0.991, 0.9957))
+ Boosting (gbm) model: 0.9662 (95% CI : (0.9607, 0.971))
+ Decision Tree: 0.4953 (95% CI: (0.4812, 0.5094)) 

Accuracy for Random Forest & Boosting model is more than 96% but performance for Boosting model is better, therefore, Boosting model is choosen. Moreover, Boosted Trees can be distributed and very fast. The accuracy of the choosen model is 0.967. The expected out-of-sample error is estimated at 0.033. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 96% on our cross-validation data, we can expect that very few, of the test samples will be missclassified.

##### Submission

``` {r}
# predict outcome levels on the original Testing data set using Gradient Boosting algorithm
predictfinal <- predict(modFit, testing, type="raw")
predictfinal
```

```{r}
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```

##### References

+ http://xgboost.readthedocs.org/en/latest/model.html
+ http://groupware.les.inf.puc-rio.br/har

